\LaTeXe\ guide for the Journal of Functional Programming

### 2019-05-03 papers/jfp/3-abstractions.tex
Work on S8.3

But the task description cannot possibly work for \emph{any} \hs{f} whatsoever; most task
descriptions (e.g. \hs{sprsh1} in \S\ref{sec-task}) require that \hs{f}
satisfies certain properties, such as \hs{Applicative} or \hs{Monad}. That is
why \hs{Task} has the ``\hs{c}~\hs{f}~\hs{=>}'' constraint in its type,
expressing that \hs{f} can only be instantiated by types that satisfy the
constraint \hs{c} and, in exchange, the task has access to the operations of class \hs{c}.
So the type \hs{Task} emerges naturally, almost inevitably.
But now that it \emph{has} emerged, we find that constraints \hs{c} classify
task descriptions in a very interesting, and practically useful, way:
\begin{itemize}
\item \hs{Task}~\hs{Applicative}. In \hs{sprsh1} we needed only \hs{Applicative}
  operations, expressing the fact that the dependencies between cells can be
  determined \emph{statically}; that is, by looking at the formulae, without
  ``computing'' them (see \S\ref{sec-deps}).
\item \hs{Task}~\hs{Monad}. As we shall see in \S\ref{sec-task-monad}, a monadic
  task allows \emph{dynamic} dependencies, in which a formula may depend on cell
  \cmd{C}, but \emph{which} cell \cmd{C} depends on the value of another cell
  \cmd{D}.
\item \hs{Task}~\hs{Functor} is somewhat degenerate: a functorial task
  description cannot even use the application operator \hs{<*>}, which limits
  dependencies to a linear chain, as e.g. in Docker containers~\cite{docker}. It
  is interesting to note that, when run on such a task description, the
  \hs{busy} build system will build each key at most once, thus partially
  fulfilling the minimality requirement~\ref{def-minimal}. Alas, it still has no
  mechanism to decide which input keys changed since the previous build.
\item \hs{Task}~\hs{MonadFail} correspond to monadic tasks that may fail. For
  example, the formula \cmd{A1/A2} may fail due to division by zero. We will
  discuss this in~\S\ref{sec-failures}.
\item \hs{Task}~\hs{Alternative}, \hs{Task}~\hs{MonadPlus} and their
  variants can be used for describing tasks with a certain type of
  non-determinism, as discussed in~\S\ref{sec-non-determinism}.
\item \hs{Task}~\hs{(MonadState}~\hs{i)} will be used
  in~\S\ref{sec-implementations} to describe tasks that have read and write
  access to the persistently stored build information~\hs{i}.
\end{itemize}

>>>>

But the task description cannot possibly work for \emph{any} \hs{f} whatsoever; most task
descriptions (e.g. \hs{sprsh1} in \S\ref{sec-task}) require that \hs{f}
satisfies certain properties, such as \hs{Applicative} or \hs{Monad}. That is
why \hs{Task} has the ``\hs{c}~\hs{f}~\hs{=>}'' constraint in its type,
expressing that \hs{f} can only be instantiated by types that satisfy the
constraint \hs{c} and, in exchange, the task has access to the operations of class \hs{c}.
So the type \hs{Task} emerges naturally, almost inevitably.
But now that it \emph{has} emerged, we find that constraints \hs{c} classify
task descriptions in a very interesting, and practically useful, way:
\begin{itemize}
\item \hs{Task}~\hs{Applicative}. In \hs{sprsh1} we needed only \hs{Applicative}
  operations, expressing the fact that the dependencies between cells can be
  determined \emph{statically}; that is, by looking at the formulae, without
  ``computing'' them (see \S\ref{sec-deps}).
\item \hs{Task}~\hs{Monad}. As we shall see in \S\ref{sec-task-monad}, a monadic
  task allows \emph{dynamic} dependencies, in which a formula may depend on cell
  \cmd{C}, but \emph{which} cell \cmd{C} depends on the value of another cell
  \cmd{D}.
\item \hs{Task}~\hs{Functor} is somewhat degenerate: a functorial task
  description cannot even use the application operator \hs{<*>}, which limits
  dependencies to a linear chain, as e.g. in Docker containers~\cite{docker}. It
  is interesting to note that, when run on such a task description, the
  \hs{busy} build system will build each key at most once, thus partially
  fulfilling the minimality requirement~\ref{def-minimal}. Alas, it still has no
  mechanism to decide which input keys changed since the previous build.
\item \hs{Task}~\hs{MonadFail} correspond to monadic tasks that may fail. For
  example, the formula \cmd{A1/A2} may fail due to division by zero. We will
  discuss this in~\S\ref{sec-failures}.
\item \hs{Task}~\hs{MonadPlus}, \hs{Task}~\hs{MonadRandon} and their
  variants can be used for describing tasks with a certain type of
  non-determinism, as discussed in~\S\ref{sec-non-determinism}.
\item \hs{Task}~\hs{(MonadState}~\hs{i)} will be used
  in~\S\ref{sec-implementations} to describe tasks that have read and write
  access to the persistently stored build information~\hs{i}.
\end{itemize}

### 2019-05-03 papers/jfp/8-engineering.tex
Work on S8.3

Most of the implementations in \S\ref{sec-implementations} can deal with
non-determinism, apart from \Buck, which relies on deterministic tasks, and
in turn can optimise the number of roundtrips required to the server.

@@

\noindent
Such tasks can be modelled in our framework by adjusting the correctness
definition~(\S\ref{sec-build-correctness}), where instead of requiring that the
\hs{result}ing value \emph{equals} the one obtained by recomputing the \hs{task}:
\[
\hs{getValue}~\hs{k}~\hs{result}~\hs{==}~\hs{compute}~\hs{task}~\hs{result} %\text{,}
\]
\noindent
we now require that \hs{result} contains \emph{one possible result of
recomputing} the \hs{task}:
\[
\hs{getValue}~\hs{k}~\hs{result}~\hs{`elem`}~\hs{computeND}~\hs{task}~\hs{result} %\text{,}
\]
where
\hs{computeND}~\hs{::}~\hs{Task}~\hs{MonadPlus}~\hs{k}~\hs{v}~\hs{->}~\hs{Store}~\hs{i}~\hs{k}~\hs{v}~\hs{->}~\hs{[@@v]}
returns the list of all possible results of the \hs{task} instead of just one
value (`ND' stands for `non-deterministic').

@@

Note that \hs{Task}~\hs{MonadPlus} is powerful enough to model dependency-level
non-determinism, for example, \cmd{INDIRECT("A" \& RANDBETWEEN(1,2))}, whereas
most build tasks in real-life build systems only experience a value-level
non-determinism. \Excel handles this example simply by marking the cell
volatile~--~an approach that can be readily adopted by any of our
implementations.

>>>>

Most of the implementations in~\S\ref{sec-implementations} can deal with
non-determinism, apart from \Buck, which uses the assumption of task determinism
to reduce the number of cloud lookups: if all tasks are deterministic, one needs
just a single cloud lookup for obtaining the value of the target key from the
hash of its terminal inputs. This is highly desirable not only for shallow
builds, but for non-shallow builds too, because under the assumption of task
determinism \emph{all} intermediate values are fully determined by the terminal
inputs, and can therefore be requested in a single batch query to the server.

@@

\noindent
Such tasks can be modelled in our framework by adjusting the correctness
definition~(\S\ref{sec-build-correctness}): instead of requiring that the
resulting value \emph{equals the result} of recomputing the task, we now require
that resulting value \emph{belongs to the set of possible results} of
recomputing the task, i.e. the set \cmd{$\{$A1 + 1, A1 + 2$\}$} in the above
example.

@@

Note that \hs{Task}~\hs{MonadRandom} is powerful enough to model
dependency-level non-determinism, for example,
\cmd{INDIRECT("A" \& RANDBETWEEN(1,2))}, whereas most build tasks in real-life
build systems only experience a value-level non-determinism. \Excel handles this
example simply by marking the cell volatile~--~an approach that can be readily
adopted by any of our implementations.
